First Order Alternation

### 2018-10-18 quantifiers/draft.tex
ok

A function $\xi : \nat \rightarrow \nat$ is [strictly]
\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
     finite-range, we denote by $\xi^{-1}(n) \in \nat$ the maximal
     value $m$ such that $\xi(m)=n$. It is easy to see that $\xi^{-1}$
     is strictly monotonic if $\xi$ is monotonic and finite-range. The
     lemma below gives the proof of correctness for the previous
     construction of non-local GLI.

@@

  \noindent(\ref{it2:subst-quant}) With the notation introduced at
  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
  \tau_{i_\ell}$, whose existence is proved by point
  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
  strictly monotonic, we obtain: \begin{compactitem}
  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
    \stamp{X}{0} \cup \set{x_j \mid j < \xi^{-1}(0), Q_j = \exists}$,
    %
  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
    %
  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
  \end{compactitem}
  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
        \max(\xi^{-1}(0)), Q_j = \exists}$,
      %
    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
      %
    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
  \end{compactitem}
  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
      Theorem \ref{thm:skolem},
      %
    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
      I_{k-1} \wedge \phi_k \models I_k$, and 
      %
    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
      \models I_n \wedge \psi \models \bot$,     
  \end{compactitem}
  as required by Definition
  \ref{def:generalized-lyndon-interpolant}. \qed}

In conclusion, under two assumptions about the first-order theory of
the data domain, namely\begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \max(\xi^{-1}(k)), Q_j = \exists}$. The remaining question is how to
use such non-local interpolants to label the unfolding of an automaton
(Definition \ref{def:unfolding}) and to compute the coverage between
nodes of the unfolding (Definition \ref{def:coverage}).

@@

As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
other than predicate symbols, such that the labeling is compatible
with the transition relation of the automaton, according to the point
(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. Suppose that
a given input event sequence $\alpha \in \dom(U)$ is labeled with a
sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1), \ldots
U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement from
Definition \ref{def:unfolding} (\ref{it3:unfolding}) and, moreover, it
has a non-local GLI $(I_0, \ldots, I_n)$, such that $\voc{I_k}
\subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
renamed variables occurring in the rules of $\Delta$ under the scope
of existential transition quantifiers\footnote{These are the
  existentially quantified variables from the prenex normal form of
  $\quantform{\alpha_k}$.}. The interpolant is used to refine the
labeling of the sequence as follows, for all $k \in
[n]$: \begin{compactitem}
\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
  from $I_k$ by \begin{inparaenum}[(i)]
    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
      %
    \item existentially quantifying each variable $y \in
      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$,
  \end{inparaenum}
%
\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
\end{compactitem}

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data
values seen in the input at some earlier moment in time, or 
%
\item variables $\set{x_{i_1},\ldots,x_{i_k}}$ that track past choices
  made within the transition rules. 
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, as only the value
of the variable determines the future behavior. Intuitively,
existential quantification of these variables does the job of ignoring
when in the past these values have been seen. Formally, the new set of
formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
\ref{def:unfolding}, because: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, since, by point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
  \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
  U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
%
\item by Definition \ref{def:generalized-lyndon-interpolant}
  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
  [n-1]$:
  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
  \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] Then the following
  entailment holds:
  \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
  Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
  satisfies the labeling condition of Definition \ref{def:unfolding}
  (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
  we obtain, as required: 
  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models
  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
\end{compactitem}

The last ingredient of the lazy annotation semi-algorithm based on
unfolding refinement with coverage consist in the implementation of
the coverage check, when the unfolding of an automaton is labeled with
conjunctions of existentially quantified formulae with predicate
symbols, obtained from interpolation. By Definition
\ref{def:coverage}, checking whether a given node $\alpha \in \dom(U)$
is covered amounts to finding a prefix $\alpha' \prefix \alpha$ and a
node $\beta \in \dom(U)$ such that $U(\alpha') \models U(\beta)$, or
equivalently, the formula $U(\alpha') \wedge \neg U(\beta)$ is
unsatisfiable. However, the latter formula, in prenex form, has
quantifier prefix in the language $\exists^*\forall^*$ and, as
previously mentioned, the satisfiability problem for such formulae
becomes undecidable when the data theory subsumes Presburger
arithmetic \cite{Halpern91}.

Nevertheless, if we require just a yes/no answer (i.e.\ not an
interpolant) recently developped quantifier instantiation heuristics
\cite{ReynoldsKK17} perform rather well in answering a large number of
queries in this class. Observe, moreover, that coverage does not need
to rely on a complete decision procedure. If the prover fails in
answering the above satisfiability query, then the semi-algorithm
assumes the node is not covered and continues exploring its
successors. Failure to compute complete coverage may lead to
divergence (non-termination) and failure to prove emptiness, but does
not affect the soundness of the semi-algorithm (real counterexamples
will still be found).

@@

The main application of first-order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, we can extend our previous
\emph{generic register automata} \cite{HolikIosifRogalewiczVojnar18}
inclusion checking framework, by allowing monitor (right-hand side)
automata to have local variables, that are not visible in the
language\footnote{We conjecture that hidden computations using
  invisible variables (clocks) are the reason for which timed and
  register automata cannot be complemented within their class.}.

@@

A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
where: $\Sigma$ is a finite set of input events, $S$ is a finite set
of states, $S_0, F \subseteq S$ are sets of initial and final states,
respectively, $C$ is a finite set of clocks and $E \subseteq S \times
\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
\item $s_0 \in S_0$ and $\gamma_0(x)$ for all $x \in C$, 
%
\item for all $i \in [n]$, there exists a transition
  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
\end{compactitem}
The run is accepting iff $s_n \in F$, in which case $T$ accepts
$w$. As usual, we denote by $\lang{T}$ the set of finite words
accepted by $T$. It is well-known \cite{AlurDill94} that, in general,
there is no timed automaton accepting the complement language
$(\Sigma\times\real)^* \setminus \lang{T}$ and, moreover, the language
inclusion problem is undecidable.

Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
first-order alternating automaton $\A_T =
\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
variable $t$, ranging over $\real$, such that each timed word $w =
(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
  w\in\lang{T}}$. The only difficulty here is capturing the fact that
all the clocks of $T$ evolve at the same pace, which is easily done
using a technique from \cite{Fribourg98}, that consists in replacing
each clock $x_i$ of $T$ by a variable $y_i$ who tracks the difference
between the values of $t$ and $x_i$, instead of the value of $x_i$
directly.

Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
$\Delta_T$ contains the rule:
\[\begin{array}{rcl}
q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} \exists y'_1 \ldots \exists y'_k
~.~ t > z \wedge \delta(z-y_1,\ldots,z-y_k) & \wedge & 
 \bigwedge_{x_i \in \lambda} y'_i = z \wedge \bigwedge_{x_i \not\in \lambda} y'_i = y_i \\ 
& \wedge & q_j(y'_1,\ldots,y'_k,t)
\end{array}\]
Moreover, nothing else is in $\Delta_T$. We establish the following
connection between a timed automaton and its corresponding first-order
alternating automaton.

@@

Since thread identifiers can only be compared for equality, the data
theory of predicate automata is the first-order theory of
equality. Moreover, transition quantifiers are only needed for
checking termination and, generally, liveness properties
\cite{Farzan16}. Thus, syntactically speaking, predicate automata are
a subset of first-order alternating automata.

>>>>

A function $\xi : \nat \rightarrow \nat$ is [strictly]
\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
     maximal value $m$ such that $\xi(m)=n$. It is easy to see that
     $\xi_{\max}^{-1}$ is strictly monotonic if $\xi$ is monotonic and
     finite-range. The lemma below gives the proof of correctness for
     the previous construction of non-local GLI.

@@

  \noindent(\ref{it2:subst-quant}) With the notation introduced at
  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
  \tau_{i_\ell}$, whose existence is proved by point
  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
  strictly monotonic, we obtain: \begin{compactitem}
  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
    \stamp{X}{0} \cup \set{x_j \mid j < \xi_{\max}^{-1}(0), Q_j = \exists}$,
    %
  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
      \xi_{\max}^{-1}(i), Q_j = \exists}$, for all $i \in [1,n]$,
    %
  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
  \end{compactitem}
  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
        \xi_{\max}^{-1}(0), Q_j = \exists}$,
      %
    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
       \xi_{\max}^{-1}(k), Q_j = \exists}$, for all $k \in [1,n]$,
      %
    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
  \end{compactitem}
  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
      Theorem \ref{thm:skolem},
      %
    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
      I_{k-1} \wedge \phi_k \models I_k$, and 
      %
    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
      \models I_n \wedge \psi \models \bot$,     
  \end{compactitem}
  as required by Definition
  \ref{def:generalized-lyndon-interpolant}. \qed}

In conclusion, under two assumptions about the first-order theory of
the data domain, namely the \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
to use such non-local interpolants to label the unfolding of an
automaton (Definition \ref{def:unfolding}) and to compute the coverage
between nodes of the unfolding (Definition \ref{def:coverage}).

@@

As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
other than predicate symbols, such that the labeling is compatible
with the transition relation of the automaton, according to the point
(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. The following
lemma describes the refinement of the labeling of an input sequence
$\alpha$ of length $n$ by a non-local GLI $(I_0, \ldots, I_n)$, such
that $\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\vec{x}_k$, where $\vec{x}_k$ are the existentially quantified
variables from the prenex normal form of $\quantform{\alpha_k}$.

\begin{lemma}\label{lemma:refinement}
  Let $U$ be an unfolding of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
  \in \dom(U)$ and $(I_0, \ldots, I_n)$ be a GLI for $\alpha$. The
  mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ defined as: 
  \begin{compactitem}
  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
    where $J_k$ is the formula obtained from $I_k$ by replacing each
    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
    existentially quantifying each free variable in $I_k$,
  %
\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
  \not\prefix \alpha$,
  \end{compactitem}
  is an unfolding of $\A$. 
\end{lemma}
\proof{
  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
  with Definition \ref{def:unfolding}, because: \begin{compactitem}
  \item $U'(\alpha_0) \equiv \iota$, since, by point
    \ref{it2:generalized-lyndon-interpolant} of Definition
    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
    %
  \item by Definition \ref{def:generalized-lyndon-interpolant}
    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
    [n-1]$:
    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
    $\overstamp{I}{j}_k$ for the formula in which each predicate
    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
    following entailment holds:
    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
    existentially quantifying all the free variables of $I_k$, we also
    obtain, applying Fact \ref{fact:ex-entail} below:
    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
    satisfies the labeling condition of Definition \ref{def:unfolding}
    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
    we obtain, as required: 
    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models
    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
  \end{compactitem}

  \begin{fact}\label{fact:ex-entail}
    Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
    that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
    $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
    ~.~ \psi(\vec{x})$. 
  \end{fact}
  \proof{ For each choice of a valuation for the existentially
    quantified variables on the left-hand side, we chose the same
    valuation for the variables on the right-hand side. \qed}\qed}

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
  in the input at some earlier moment in time, and
%
\item variables that track past choices made within the transition
  rules.
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, as only the value
of the variable determines the future behavior. Intuitively,
existential quantification of these variables does the job of ignoring
when in the past these values have been seen.

The last ingredient of the lazy annotation semi-algorithm based on
unfoldings consist in the implementation of the coverage check, when
the unfolding of an automaton is labeled with conjunctions of
existentially quantified formulae with predicate symbols, obtained
from interpolation. By Definition \ref{def:coverage}, checking whether
a given node $\alpha \in \dom(U)$ is covered amounts to finding a
prefix $\alpha' \prefix \alpha$ and a node $\beta \in \dom(U)$ such
that $U(\alpha') \models U(\beta)$, or equivalently, the formula
$U(\alpha') \wedge \neg U(\beta)$ is unsatisfiable. However, the
latter formula, in prenex form, has quantifier prefix in the language
$\exists^*\forall^*$ and, as previously mentioned, the satisfiability
problem for such formulae becomes undecidable when the data theory
subsumes Presburger arithmetic \cite{Halpern91}.

Nevertheless, if we require just a yes/no answer (i.e.\ not an
interpolant) recently developped quantifier instantiation heuristics
\cite{ReynoldsKK17} perform rather well in answering a large number of
queries in this class. Observe, moreover, that coverage does not need
to rely on a complete decision procedure. If the prover fails in
answering the above satisfiability query, then the semi-algorithm
assumes that the node is not covered and continues exploring its
successors. Failure to compute complete coverage may lead to
divergence (non-termination) and ultimately, to failure to prove
emptiness, but does not affect the soundness of the semi-algorithm
(real counterexamples will still be found).

@@

The main application of first-order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, the method described in
\S\ref{sec:emptiness} can extend our previous \emph{generic register
  automata} \cite{HolikIosifRogalewiczVojnar18} inclusion checking
framework, by allowing monitor (right-hand side) automata to have
local variables, that are not visible in the language\footnote{We
  conjecture that hidden computations using invisible variables
  (clocks) are the reason for which timed and register automata cannot
  be complemented within their class.}.

@@

A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
where: $\Sigma$ is a finite set of input events, $S$ is a finite set
of states, $S_0, F \subseteq S$ are sets of initial and final states,
respectively, $C$ is a finite set of clocks and $E \subseteq S \times
\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
\item $s_0 \in S_0$ and $\gamma_0(x)=0$ for all $x \in C$, 
%
\item for all $i \in [n]$, there exists a transition
  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
\end{compactitem}
The run is accepting iff $s_n \in F$, in which case $T$ accepts
$w$. As usual, we denote by $\lang{T}$ the set of finite words
accepted by $T$. It is well-known that, in general, there is no timed
automaton accepting the complement language $(\Sigma\times\real)^*
\setminus \lang{T}$ and, moreover, the language inclusion problem is
undecidable \cite{AlurDill94}.

Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
first-order alternating automaton $\A_T =
\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
variable $t$, ranging over $\real$, such that each timed word $w =
(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
  w\in\lang{T}}$. The only difficulty here is capturing the fact that
all the clocks of $T$ evolve at the same pace, which is easily done
using a technique from \cite{Fribourg98}, which replaces
each clock $x_i$ of $T$ by a variable $y_i$ tracking the difference
between the values of $t$ and $x_i$.

Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
$\Delta_T$ contains the rule:
\[q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} t > z \wedge \delta(z-y_1,\ldots,z-y_k) \wedge q_j(y'_1,\ldots,y'_k,t)\]
where $y'_i$ stands for $z$ if $x_i \in \lambda$ and for $y_i$, 
otherwise.  Moreover, nothing else is in $\Delta_T$. We establish the
following connection between a timed automaton and its corresponding
first-order alternating automaton.

@@

Since thread identifiers can only be compared for equality, the data
theory of predicate automata is the first-order theory of
equality. Moreover, transition quantifiers are only needed for
checking termination and, generally, liveness properties
\cite{Farzan16}.
